Evaluation of Cache-Oblivious Data Structures [draft-1]

### 2008-05-26 paper/paper.tex
Added new plots.
Some more updates to the paper.

\title{Evaluation of Cache-Oblivious Data Structures}
\subtitle{\textit{Work In Progress!}}

@@

% TODO: update keywords?
\keywords{cache efficiency, locality of reference, algorithms, empirical evaluation}

@@

\subsection{Experimental Framework}

% Describe (globally) what to measure: execution time
% - NIPS VM
% - SPIN model checker
% - Promela-to-NIPS compiler

For our experimental framework, we need a collection of formal models that are
representative of those typically used for formal verification, and a way to
execute them. For the first part, we have looked at Spin \cite{spin}, a widely
used model checking tool. Spin uses a custom modeling language, called
PROMELA, to specify models and comes with a collection of example models that
are suitable for our experiment.

@@

We now have a single operation that must be implemented by all set data structures. Note that the values that need to be stored in the state space search algorithm are virtual machine state descriptions, which (in the NIPS VM) are opaque variable-length strings. All data structures must therefore support storing strings of varying length efficiently.

@@

Note that if we insert several different values, some values may map to the
same slot, which is problematic if each slot can only store one value. There
are a lot of different ways to resolve this collision problem; we use
\emph{separate chaining} which means that slots do not store values directly,
but instead each slot stores a pointer to a singly-linked list of values that
map to that slot. This particular implementation technique is well-suited to
the scenario where values may have different sizes (as the slots only need to
store a fixed-size pointer and not a variable-size value) and maintains
relatively good performance when the number of values stored exceeds the size
of the index array \cite{sedgewick1998ac}.

@@

\subsection{Measurements}
% Describe how performance was measured
%  (mock set implementation; times/sizes subtracted)

% Describe data set used; per model:
%   average state size
%   total size
%   ratio of insertions/queries

% Note that state size depends on number of processes

The following models are used for benchmarking:
\begin{itemize}
\item\verb#eratosthenes# is a parallel implementation of the sieve
of Eratosthenes which is used to find prime numbers upto a given maximum.
It has a single parameter $N$: only integers from 2 to $N$ (exclusive) are
tested. A new process is created for every prime number found, and as a result
the state space can only by exhaustively searched for relatively small values
of $N$ (e.g. $N < 40$). Because processes are dynamically created, state size
also varies during execution.

@@

\item\verb#peterson_N# is a Peterson's solution to the multi-process mutual
exclusion problem (using global variables instead of channels for
communication) with a configurable number of processes ($N$). The processes
are created atomically, so the state size is constant after initialization.

% mobile1
% pftp
% snoopy
% sort

% TODO: table of models with parameters,

\begin{table}
\begin{center}
\begin{tabular}{ l l l l }
\hline
\textbf{Case} & \textbf{Parameters} & \textbf{Iterations} & \textbf{Transitions} \\
\hline
eratosthenes & $N=40$ & 1,019,960 & 4,923,218 \\
leader2 & $N=5$ & 5,950,945 & 23,856,363 \\
peterson\_N & \\
\hline
\end{tabular}
\caption{Properties of test cases used}
\label{tab-cases}
\end{center}
\end{table}

\end{itemize}

@@

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-bender_wctime}
\caption{Eratosthenes/Bender set execution time}
\label{fig-eratosthenes-bender_wctime}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-bender_vss}
\caption{Eratosthenes/Bender set memory usage}
\label{fig-eratosthenes-bender_vss}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-hash_wctime}
\caption{Eratosthenes/hash table execution time}
\label{fig-eratosthenes-hash_wctime}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-hash_vss}
\caption{Eratosthenes/hash table memory usage}
\label{fig-eratosthenes-hash_vss}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-btree_wctime}
\caption{Eratosthenes/B-tree execution time}
\label{fig-eratosthenes-btree}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-btree_vss}
\caption{Eratosthenes/B-tree memory usage}
\label{fig-eratosthenes-btree}
\end{figure}

% Present results!
% Tables, graphs, etc.
% What to include:
%   queue size over iterations
%       for each program
%
%   insertions per query:
%       for each program
%
%   per program:
%       wctime over iterations
%          for each set
%       memory over iterations
%          for each set
%

\section{Discussion}
% Explain results, draw conclusions.

\section{Conclusion}
% Repeat findings.

\include{fig-queuesize}

>>>>

\title{Evaluation of Cache-Oblivious Data Structures [draft-1]}

@@

\keywords{cache efficiency, locality of reference, algorithms, empirical evaluation}

@@

\subsection{Experimental Framework}
For our experimental framework, we need a collection of formal models that are
representative of those typically used for formal verification, and a way to
execute them. For the first part, we have looked at Spin \cite{spin}, a widely
used model checking tool. Spin uses a custom modeling language, called
PROMELA, to specify models and comes with a collection of example models that
are suitable for our experiment.

@@

We now have a single operation that must be implemented by all set data structures. Recall that the values to be stored are virtual machine state descriptions, which are variable-length binary strings. All data structures must therefore support storing strings of varying length efficiently.

@@

When several values are inserted, some values may map to the same slot,
which is problematic if each slot can only store one value. There
are a lot of different ways to resolve this collision problem; we use
\emph{separate chaining} which means that slots do not store values directly,
but instead each slot stores a pointer to a singly-linked list of values that
map to that slot. This particular implementation technique is well-suited to
the scenario where values may have different sizes (as the slots only need to
store a fixed-size pointer and not a variable-size value) and maintains
relatively good performance when the number of values stored exceeds the size
of the index array \cite{sedgewick1998ac}.

@@

\subsection{Measurements}
Measuring the memory usage of a process can be done by retrieving how much of the
address space has been allocated by a process; typically called the
\emph{virtual set size}.
Since a single page can be mapped into memory more than once (either in multiple
processes, or in the same process) this is not exactly equal to the amount of
memory allocated for the process, but since the majority of the memory used by
the data structures is mapped at exactly one location, it is good enough as
an indication of memory use.

There are several ways of measuring the time a process takes to execute.
The simplest is counting the number of seconds elapsed since the start of the
program (called "wall clock time"). This has the disadvantage that concurrently
executing processes affect the runtime of the process being measured, so on a
busy system the reported runtime may vary with different executions.

A different way to measure time is using the statistics the kernel collects
of how many seconds the CPU spends executing instructions for the process
(both in user space and in kernel space). These values are independent of
what other processes are doing, which makes them more consistent across
multiple executions. However, these values do not account for the time the
system spends waiting (for example, for a page to be swapped in from disk)
or the time spend by other processes on behalf of the executing process
(for example, the kernel swap daemon on Linux). Since these factors may have
a large effect on the total running time of the algorithm, it is undesirable
to leave these out.

Therefore, we decided to measure wall clock time. In order to deal with
variations across multiple executions, each experiment is run three times,
and the median value is used for further analysis.

\subsubsection{Framework Overhead}
Finally, there is another factor that must be taken into account: the test
framework uses several components (such as the NIPS VM and a queue) that are
not being evaluated, yet which do contribute to the memory use and runtime
of the process. In order to discount these factors, tests are performed using
a mock set implementation that gives a near-minimal overhead. This works
by first running with a real set implementation and logging all results
(the return values of the \verb#insert(S,x)# calls) to a file, and then
running a second time while reading the stored values. As a result, the mock
set implementation has very little memory and runtime overhead.

In the results below, all values are reported relative to the values obtained
using the mock set implementation, which means the overhead of the test
framework is removed from the results.

\subsubsection{System Configuration}
The experiments where performed on a 64-bit Linux (kernel 2.6) system
with an Intel Xeon E5320 processor (1.86 GHz, 4MB cache) and 8GB of main
memory and 2GB swap file. Although the system is a multiprocessor system
(and the E5320 is a dual-core processor) all code is single-threaded so
only a single core is used when executing tests.

% Describe data set used; per model:
%   average state size?
%   total size?
%   ratio of insertions/queries?

The following models are used for benchmarking:
\begin{itemize}
\item\verb#eratosthenes# is a parallel implementation of the sieve
of Eratosthenes which is used to find prime numbers upto a given maximum.
It has a single parameter $N$: only integers from 2 to $N$ (exclusive) are
tested. A new process is created for every prime number found, and as a result
the state space can only by exhaustively searched for relatively small values
of $N$ (e.g. $N < 40$). Because processes are dynamically created, state size
increases during execution.

@@

\item\verb#peterson_N# is a Peterson's solution to the multi-process mutual
exclusion problem (using global variables instead of channels for
communication) with a configurable number of processes ($N$). The processes
are created atomically, so the state size is constant after initialization.
\end{itemize}

\begin{table}
\begin{center}
\begin{tabular}{ l l l l }
\hline
\textbf{Case} & \textbf{Parameters} & \textbf{Iterations} & \textbf{Transitions} \\
\hline
eratosthenes & $N=40$ &  1,019,960 &  4,923,218 \\
leader2      & $N=5$  &  5,950,945 & 23,856,363 \\
peterson\_N  & $N=4$  & 10,000,000 & 37,434,411 \\
\hline
\end{tabular}
\caption{Properties of test cases used}
\label{tab-cases}
\end{center}
\end{table}

In Table \ref{tab-cases} an overview is given of the parameters of the models used
and the properties of the execution of the model. Note that the number of transitions
relative to the number of iterations indicates the percentage of insertion operations
that results in a new element being added to the set.

% Possible additional cases:
% - mobile1
% - pftp
% - snoopy
% - sort

@@

TODO: describe parameters of data structures
Explain results for B-tree (higher page size packs more data), hash table (larger index, less collissions, cache miss anyway)

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-queuesize}
\caption{Queue size}
\label{fig-quesize}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-transitions}
\caption{Transitions}
\label{fig-transitions}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-bender_wctime}
\caption{Bender set execution time on Eratosthenes}
\label{fig-eratosthenes-bender_wctime}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-bender_vss}
\caption{Bender set memory usage on Eratosthenes}
\label{fig-eratosthenes-bender_vss}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-hash_wctime}
\caption{Hash table execution time on Eratosthenes}
\label{fig-eratosthenes-hash_wctime}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-hash_vss}
\caption{Hash table memory usage on Eratosthenes}
\label{fig-eratosthenes-hash_vss}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-btree_wctime}
\caption{B-tree execution time on Eratosthenes}
\label{fig-eratosthenes-btree_wctime}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-btree_vss}
\caption{B-tree memory usage on Eratosthenes}
\label{fig-eratosthenes-btree_vss}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-leader2_wctime}
\caption{Execution times on Leader2}
\label{fig-leader2_wctime}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-leader2_vss}
\caption{Memory usage on Leader2}
\label{leader2_vss}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-peterson_wctime}
\caption{Execution time on Peterson\_N}
\label{fig-peterson_wctime}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-peterson_vss}
\caption{Memory usage on Peterson\_N}
\label{fig-peterson_vss}
\end{figure}

% Present results!
% Tables, graphs, etc.
% What to include:
%   queue size over iterations
%       for each program
%
%   insertions per query:
%       for each program
%
%   per program:
%       wctime over iterations
%          for each set
%       memory over iterations
%          for each set

\section{Discussion}
% Explain results, draw conclusions.
% - relate to insert/test ratio
% - relate to hardware (not many levels of hierarchy)

\section{Conclusion}
The experimental results clearly show that in terms of both memory usage and execution time,
the cache-oblivious data structure proposed by Bender, Duan, Iacono and Wu is outperformed
by traditional data structures in our test scenario.
The advantage of more cache-friendly behaviour does not appear to be large enough to
compensate for the increased complexity of the data structure, which results in higher
memory requirements and increased computational overhead.
If the data structure does have asymptotic performance benefits, then realistic work loads
on current hardware systems are not enough to reveal this.

Of course, this does not prove definitely that cache-oblivious data structures
are entirely without merit. It should be noted that the experiments performed
for this research are limited in scope: only a single cache-oblivious data
structure has been examined, in a single test scenario, on a single hardware
and software platform. It is possible that in a different setting,
cache-oblivious data structure do compare favorably to their traditional
counterparts.

We hope that future research will focus not only on the theoretical performance
of newly developed algorithms and data structure, but also on the performance
in practice. Theoretical results are invalueble, but to be of greatest
significance newly developed data structures and algorithms should have
demonstrable practical merit as well.
